---
title: "Like Tennis's One-Handed Backhand? It Could Be Gone By 2030"
author: ''
date: '2019-05-04'
slug: tennis-one-handed-backhand
categories: []
tags: []
---


```{r results='asis', echo=FALSE, include=FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(lubridate)
library(hrbrthemes)
library(ggthemes)
library(scales)
library(plotly)
theme_set(theme_ipsum_rc() +
          theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
          theme(axis.title.y = element_text(angle = 0)))
load("C:/Users/nyulo/Documents/HBS/LM/tennis_rankings/data/atp_top_100.R")
```


## About The Dataset

I scraped ranking data from the [ATP World Tour](https://www.atptour.com/en/rankings/singles) website between the years of 1980 through present day. While I was delighted to see that the player names / rank information was complete, there are a substantial amount of missing values describing what type of backhand a player hit.

As we might expect, it turns out that the older data contains the vast majority of the missings:

```{r results='asis', echo= TRUE}

p2 <- clean %>% 
  ggplot(aes(year, fill = is.na(one_hander))) +
  geom_density(alpha = .6) +
  labs(title = "Distribution of Missing Backhand Type by Year",
       fill = "Missing?")

ggplotly(p2)
```

Perhaps more problematic (but again expected), the missing backhand type's are correlated with rank; higher ranked players were less likely to be missing backhand type:

```{r}

p3 <- clean %>% 
  mutate(Ranking = parse_number(Ranking)) %>% 
  group_by(Ranking) %>% 
  summarize(missing_share = sum(is.na(one_hander))) %>% 
  ggplot(aes(Ranking, missing_share)) +
  geom_line(size = 1.7) +
  expand_limits(y = 0) +
  labs(title = "Sum of Missing Counts By Ranking")

ggplotly(p3)
```

# ONE HANDED BACKHAND SHARE BY YEAR

Because of these shortcomings I elected to drop rankings prior to 2000. 

```{r, message = FALSE}
p4 <- clean %>% 
  filter(year >= 2000) %>% 
  group_by(date) %>% 
  summarize(share = mean(one_hander, na.rm = TRUE)) %>% 
  ggplot(aes(date, share)) +
  geom_smooth(size = 1.7) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  labs(x = "", y = "", color = "", title = "Average Share of One Handers by Year")

ggplotly(p4)
```

# ONE HANDED BACKHAND RATE: TOP 50 VS 51 - 100

```{r}
p5 <- clean %>% 
  filter(year >= 2000) %>% 
  group_by(year, rank_cohort) %>% 
  summarize(share = mean(one_hander, na.rm = TRUE)) %>% 
  ggplot(aes(year, share, color = as.factor(rank_cohort))) +
  geom_smooth(se = FALSE, size = 1.7) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  scale_y_percent() +
  labs(x = "", y = "", color = "", title = "Percent of One Handed Backhands in ATP Top 100")

ggplotly(p5)
```

# Again

```{r results='asis', echo=FALSE}
# ONE HANDED BACKHAND - EXTENDED MODEL
p6 <- clean %>% 
  filter(year >= 2000) %>% 
  group_by(year) %>% 
  summarize(share = mean(one_hander, na.rm = TRUE)) %>% 
  ggplot(aes(year, share)) +
  geom_point(size = 2) +
  stat_smooth(method = "lm", fullrange = TRUE) +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) +
  expand_limits(y = 0) +
  xlim(c(2000,2030)) +
  geom_hline(yintercept = 0, linetype = 2) +
  labs(x = "", y = "", color = "", title = "Average Share of One-Handers in ATP Top 100 by Year")

ggplotly(p6)
```